### 2021-09-03 ms/ms.Rmd
biorxiv submission

* Developmental integration can cause traits to covary over macroevolutionary time and in some cases prevent populations from reaching their adaptive optima. Developmental integration between stomatal size and density may contribute to two major features of stomatal anatomical evolution: inverse size-density scaling and bimodal stomatal ratio. If these patterns result from developmental integration, we predicted that in amphistomatous leaves 1) stomatal size and density should covary similarly on both abaxial and adaxial surfaces and 2) stomatal traits (size and density) on each surface should covary isometrically.
* We synthesized data on stomatal density and length from amphistomatous leaves of `r n_tax3` terrestrial seed plant taxa mostly from the literature. We estimated the covariance in divergence between stomatal traits from `r n_pair` phylogenetically independent contrasts using a robust Bayesian model.
*	Adaxial stomatal density, but not length, is evolutionary labile and not strongly integrated with stomatal length or abaxial stomatal density. Hence, developmental integration alone cannot explain inverse size-density scaling nor bimodal stomatal  ratio.
* Quasi-independent evolution of stomatal anatomical traits facilitates largely unfettered access to fitness optima. If stomatal anatomical traits are near their current fitness optimum, this implies that limits on trait (co)variance result from selective rather than developmental constraints. However, we cannot rule out that developmental integration is important in some lineages. Future research should identify the mechanistic basis of (dis)integration in stomatal development.

@@

Two salient features of stomatal anatomy have been recognized for decades but we do not yet understand the evolutionary forces that generate and maintain them. We denote these two features as “inverse size-density scaling” and “bimodal stomatal ratio”  (Fig. \ref{fig:concepts}). Inverse size-density scaling refers to the negative interspecific correlation between the size of the stomatal apparatus and the density of stomata [@weiss_untersuchungen_1865; @franks_maximum_2009; @de_boer_optimal_2016; @sack_developmental_2016; @liu_scaling_2021]. Across species, leaves with smaller stomata tend to pack them more densely, but there is significant variation about this general trend. Bimodal stomatal ratio refers to the observation that the ratio of stomatal density on the adaxial (upper) surface to the density on the abaxial (lower) has distinct modes (Fig. 1). Stomata are most often found only on the lower leaf surface (hypostomy), but occur on both surfaces (amphistomy) in some species [@metcalfe_anatomy_1950; @parkhurst_adaptive_1978; @mott_adaptive_1982], especially herbaceous plants from open, high light habitats [@salisbury_i_1928; @mott_adaptive_1982; @gibson_structure-function_1996; @smith_associations_1998; @jordan_using_2014; @muir_making_2015; @bucher_stomatal_2017; @muir_light_2018]. @muir_making_2015 described bimodal stomatal ratio formally but the pattern is apparent in earlier comparative studies of the British flora (*cf* [@peat_comparative_1994] Fig 1). For both features, we limit our focus in this study to interspecific variation in mean trait values and do not seek to understand intraspecifc variation.

@@

Unless otherwise mentioned, we performed all data wrangling and statistical analyses in *R* version `r as.character(getRversion())` [@r_core_team_r:_2021]. The source code is available to reviewers. All source code will be publically available on GitHub upon publication (https://github.com/cdmuir/stomata-independence) and archived on Zenodo.

@@

Data on stomatal anatomy are spread over a disparate literature and we have not attempted an exhaustive synthesis of amphistomatous leaf stomatal anatomy. We began our search by reviewing papers that cited key studies of amphistomy [@parkhurst_adaptive_1978; @mott_adaptive_1982; @muir_making_2015]. We supplemented these by searching Clarivate *Web of Science* for "guard cell length" because most studies that report guard cell length also report stomatal density, whereas the reverse is not true. We identified additional studies by reviewing the literature cited of papers we found and through haphazard discovery. The final data set contained `r nrow(stomata_anatomy1)` observations of stomatal density and length from `r length(unique(stomata_anatomy1$source_id))` taxa and `r length(unique(stomata_anatomy1$source))` primary studies (Table \@ref(tab:sources)). However, many of these data were excluded if taxonomic name and phylogenetic placement could not be resolved (see below). Finally, we included a previously unpublished data set of 14 amphistomatous wild tomato species (*Solanum* sect. *Lycopersicum* and sect. *Lycopersicoides*) grown in pots under outdoor summer Mediterranean conditions previously described [@muir_unpublished_2021]. We took ab- and adaxial epidermal imprints using clear nail polish of the mid-portion of the lamina away from major veins on the terminal leaflet of the youngest, fully expanded leaf from 1-5 replicates per taxon. With a brightfield light microscope, we counted stomata in three fields of view and divided by area to obtain density and measured the average guard cell length to estimate stomatal size. The data set is publicly available as an *R* package **ropenstomata** (https://github.com/cdmuir/ropenstomata). For review, we have included the current version of **ropenstomata** and the version used for publication will be deposited on Dryad and Zenodo upon publication.

@@

We extracted `r n_pair` phylogenetically independent taxon pairs (Table \@ref(tab:pair_div)). A fully resolved, bifurcating four-taxon phylogeny can have two basic topologies: $\tt{((A,B),(C,D))}$ or $\tt{((A,B),C),D))}$. Taxon pairs include all comparisons of $\tt{A}$ with $\tt{B}$ and $\tt{C}$ with $\tt{D}$ in each four-taxon clade. We extracted pairs using the `extract_sisters()` function in R package **diverge** version `r packageVersion("diverge")` [@anderson_diverge_2021] and custom scripts (see source code). Taxon pairs are the most closely related pairs in our data set, but not usually the most closely related species in nature. For each pair we calculated phylogenetically independent contrasts [@felsenstein_phylogenies_1985] as the difference in the log$_{10}$-transformed trait value. Contrasts are denoted as $\Delta \text{log(trait)}$. We log-transformed traits for normality because like many morphological and anatomical traits they are strongly right-skewed. Log-transformation also helps compare density and length, which are measured on different scales, because log-transformed values quantify proportional rather than absolute divergence. We only used pairs of terminal taxa rather than the entire tree for two reasons. First, even in whole-tree methods, approximately half of the data comes from divergence at the tips, so we do not lose much statistical power. Second, using a taxon-pair method obviates making strong assumptions about multivariate trait evolution process homogeneity throughout seed plant evolution. Homogeneity is biologically unlikely and it is difficult to fit more complex models reliably with only data from extant taxa.

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/h2.pdf}
\caption{Developmental integration in stomatal length is much stronger than stomatal density between the surfaces of amphistomatous leaves (a) Data from `r n_pair` phylogenetically independent contrasts of change in the abaxial trait value ($x$-axis) against change in the adaxial trait value ($y$-axis) for log(stomatal density) ($x$-axis) and log(stomatal density) (left panel) and log(stomatal length) (right panel). Each contrast is shown by black points and every contrast appears on both panels. Dashed grey lines are 1:1 lines for reference. Solid grey lines and ribbon the fitted SMA slope and 95\% HPD interval. (b) The SMA slope (left panel) is significantly less than 1 (isometry, top dashed line) for density but very close to isometric for length. The coefficient of determination ($r^2$, right panel) is also much greater for length than density. The points are parameter estimates with 66\% (thick lines) and 95\% HPD intervals. Reported paremeter estimates are conditioned on zero time since divergence between taxa (see \protect\hyperlink{results}{Results}).}
\label{fig:h2}
\end{figure}

@@

Developmental integration leads to trait covariation and may hinder adaptation by preventing traits from evolving independently toward a multivariate phenotypic optimum. Two major features of stomatal anatomical variation at the macroevolutionary scale, inverse size-density scaling and bimodal stomatal ratio, may be shaped by developmental integration between pavement and guard cell size on both leaf surfaces. In this study, we took advantage of the fact that amphistomatous leaves produce stomata on both abaxial (usually lower) and adaxial (usually upper) surfaces to test predictions of developmental integration hypotheses using a global phylogenetic comparison of seed plants, albeit mostly angiosperms. Contrary to developmental integration hypotheses, adaxial stomatal density ($D_\text{ad}$) evolves somewhat independently of adaxial stomatal length ($L_\text{ad}$) and abaxial stomatal density ($D_\text{ab}$). Hence, inverse size-density scaling and bimodal stomatal ratio cannot be attributed entirely to developmental integration. Quasi-independent evolution of traits should enable lineages to reach their fitness optimum, implying that selection is most likely a major constraint on the variation in stomatal anatomy in seed plants. Future research should test whether developmental integration is more or less important in certain lineages and identify the mechanistic basis of developmental disintegration where it occurs on the plant tree of life.

@@

The disintegration of stomatal size and density on adaxial leaf surfaces implies that the inverse size-density scaling on abaxial surfaces [@weiss_untersuchungen_1865; @franks_maximum_2009; @de_boer_optimal_2016; @sack_developmental_2016; @liu_scaling_2021] is not a developmental *fait accompli*. The lability of $D_\text{ad}$ may explain why there is so much putatively adaptive variation in the trait along light gradients [@muir_light_2018] and in coordination with other anatomical traits [@pathare_increased_2020]. It also suggests that the relationship between genome size and stomatal anatomy at macroevolutionary scales [@roddy_scaling_2020] may not be causal. Genome size sets a minimum on meristematic cell volume [@simova_geometrical_2012], but the decoupling of size and density on the adaxial surface suggests that meristematic cell volume is probably not a strong constraint on the final size of pavement and guard cells.

>>>>

* Developmental integration can cause traits to covary over macroevolutionary time and in some cases prevent populations from reaching their adaptive optima. Developmental integration between stomatal size and density may contribute to two major features of stomatal anatomical evolution: inverse size-density scaling and bimodal stomatal ratio. If these patterns result from developmental integration, we predicted that in amphistomatous leaves 1) stomatal size and density should covary similarly on both abaxial and adaxial surfaces and 2) stomatal traits (size and density) on each surface should covary isometrically.
* We synthesized data on stomatal density and length from amphistomatous leaves of `r n_tax3` terrestrial seed plant taxa mostly from the literature. We estimated the covariance in divergence between stomatal traits from `r n_pair` phylogenetically independent contrasts using a robust Bayesian model.
*	Adaxial stomatal density, but not length, is evolutionarily labile and not strongly integrated with stomatal length or abaxial stomatal density. Hence, developmental integration alone cannot explain inverse size-density scaling nor bimodal stomatal  ratio.
* Quasi-independent evolution of stomatal anatomical traits facilitates largely unfettered access to fitness optima. If stomatal anatomical traits are near their current fitness optimum, this implies that limits on trait (co)variance result from selective rather than developmental constraints. However, we cannot rule out that developmental integration is important in some lineages. Future research should identify the mechanistic basis of (dis)integration in stomatal development.

@@

Two salient features of stomatal anatomy have been recognized for decades but we do not yet understand the evolutionary forces that generate and maintain them. We denote these two features as “inverse size-density scaling” and “bimodal stomatal ratio”  (Fig. \ref{fig:concepts}). Inverse size-density scaling refers to the negative interspecific correlation between the size of the stomatal apparatus and the density of stomata [@weiss_untersuchungen_1865; @franks_maximum_2009; @de_boer_optimal_2016; @sack_developmental_2016; @liu_scaling_2021]. Across species, leaves with smaller stomata tend to pack them more densely, but there is significant variation about this general trend (Fig. \ref{fig:concepts}a). Bimodal stomatal ratio refers to the observation that the ratio of stomatal density on the adaxial (upper) surface to the density on the abaxial (lower) has distinct modes (Fig. \ref{fig:concepts}b). Stomata are most often found only on the lower leaf surface (hypostomy), but occur on both surfaces (amphistomy) in some species [@metcalfe_anatomy_1950; @parkhurst_adaptive_1978; @mott_adaptive_1982], especially herbaceous plants from open, high light habitats [@salisbury_i_1928; @mott_adaptive_1982; @gibson_structure-function_1996; @smith_associations_1998; @jordan_using_2014; @muir_making_2015; @bucher_stomatal_2017; @muir_light_2018]. @muir_making_2015 described bimodal stomatal ratio formally but the pattern is apparent in earlier comparative studies of the British flora (*cf.* @peat_comparative_1994 Fig 1). For both features, we limit our focus in this study to interspecific variation in mean trait values and do not seek to understand intraspecifc variation.

@@

Unless otherwise mentioned, we performed all data wrangling and statistical analyses in *R* version `r as.character(getRversion())` [@r_core_team_r:_2021]. Source code is publically available on GitHub (https://github.com/cdmuir/stomata-independence) and will be archived on Zenodo upon publication.

@@

Data on stomatal anatomy are spread over a disparate literature and we have not attempted an exhaustive synthesis of amphistomatous leaf stomatal anatomy. We began our search by reviewing papers that cited key studies of amphistomy [@parkhurst_adaptive_1978; @mott_adaptive_1982; @muir_making_2015]. We supplemented these by searching Clarivate *Web of Science* for "guard cell length" because most studies that report guard cell length also report stomatal density, whereas the reverse is not true. We identified additional studies by reviewing the literature cited of papers we found and through haphazard discovery. The final data set contained `r nrow(stomata_anatomy1)` observations of stomatal density and length from `r length(unique(stomata_anatomy1$source_id))` taxa and `r length(unique(stomata_anatomy1$source))` primary studies (Table \@ref(tab:sources)). However, many of these data were excluded if taxonomic name and phylogenetic placement could not be resolved (see below). Finally, we included some previously unpublished data. Stomatal size data were collected on grass species previously described  [@pathare_increased_2020]. We also included a previously unpublished set of 14 amphistomatous wild tomato species (*Solanum* sect. *Lycopersicum* and sect. *Lycopersicoides*) grown in pots under outdoor summer Mediterranean conditions previously described [@muir_unpublished_2021]. We took ab- and adaxial epidermal imprints using clear nail polish of the mid-portion of the lamina away from major veins on the terminal leaflet of the youngest, fully expanded leaf from 1-5 replicates per taxon. With a brightfield light microscope, we counted stomata in three fields of view and divided by area to obtain density and measured the average guard cell length to estimate stomatal size. The data set is publicly available as an *R* package **ropenstomata** (https://github.com/cdmuir/ropenstomata). It will be deposited on Dryad and archived on Zenodo upon publication.

@@

We extracted `r n_pair` phylogenetically independent taxon pairs (Table \ref{tab:pair_div}). A fully resolved, bifurcating four-taxon phylogeny can have two basic topologies: $\tt{((A,B),(C,D))}$ or $\tt{((A,B),C),D))}$. Taxon pairs include all comparisons of $\tt{A}$ with $\tt{B}$ and $\tt{C}$ with $\tt{D}$ in each four-taxon clade. We extracted pairs using the `extract_sisters()` function in R package **diverge** version `r packageVersion("diverge")` [@anderson_diverge_2021] and custom scripts (see source code). Taxon pairs are the most closely related pairs in our data set, but not usually the most closely related species in nature. For each pair we calculated phylogenetically independent contrasts [@felsenstein_phylogenies_1985] as the difference in the log$_{10}$-transformed trait value. Contrasts are denoted as $\Delta \text{log(trait)}$. We log-transformed traits for normality because like many morphological and anatomical traits they are strongly right-skewed. Log-transformation also helps compare density and length, which are measured on different scales, because log-transformed values quantify proportional rather than absolute divergence. We only used pairs of terminal taxa rather than the entire tree for two reasons. First, even in whole-tree methods, approximately half of the data comes from divergence at the tips, so we do not lose much statistical power. Second, using a taxon-pair method obviates making strong assumptions about multivariate trait evolution process homogeneity throughout seed plant evolution. Homogeneity is biologically unlikely and it is difficult to fit more complex models reliably with only data from extant taxa.

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/h2.pdf}
\caption{Developmental integration in stomatal length is much stronger than stomatal density between the surfaces of amphistomatous leaves (a) Data from `r n_pair` phylogenetically independent contrasts of change in the abaxial trait value ($x$-axis) against change in the adaxial trait value ($y$-axis) for log(stomatal density) (left panel) and log(stomatal length) (right panel). Each contrast is shown by black points and every contrast appears on both panels. Dashed grey lines are 1:1 lines for reference. Solid grey lines and ribbon the fitted SMA slope and 95\% HPD interval. (b) The SMA slope (left panel) is significantly less than 1 (isometry, top dashed line) for density but very close to isometric for length. The coefficient of determination ($r^2$, right panel) is also much greater for length than density. The points are parameter estimates with 66\% (thick lines) and 95\% HPD intervals. Reported paremeter estimates are conditioned on zero time since divergence between taxa (see \protect\hyperlink{results}{Results}).}
\label{fig:h2}
\end{figure}

@@

Developmental integration leads to trait covariation and may hinder adaptation by preventing traits from evolving independently towards a multivariate phenotypic optimum. Two major features of stomatal anatomical variation at the macroevolutionary scale, inverse size-density scaling and bimodal stomatal ratio, may be shaped by developmental integration between pavement and guard cell size on both leaf surfaces. In this study, we took advantage of the fact that amphistomatous leaves produce stomata on both abaxial (usually lower) and adaxial (usually upper) surfaces to test predictions of developmental integration hypotheses using a global phylogenetic comparison of seed plants, albeit mostly angiosperms. Contrary to developmental integration hypotheses, adaxial stomatal density ($D_\text{ad}$) evolves somewhat independently of adaxial stomatal length ($L_\text{ad}$) and abaxial stomatal density ($D_\text{ab}$). Hence, inverse size-density scaling and bimodal stomatal ratio cannot be attributed entirely to developmental integration. Quasi-independent evolution of traits should enable lineages to reach their fitness optimum, implying that selection is most likely a major constraint on the variation in stomatal anatomy in seed plants. Future research should test whether developmental integration is more or less important in certain lineages and identify the mechanistic basis of developmental disintegration where it occurs on the plant tree of life.

@@

The disintegration of stomatal size and density on adaxial leaf surfaces implies that the inverse size-density scaling on abaxial surfaces [@weiss_untersuchungen_1865; @franks_maximum_2009; @de_boer_optimal_2016; @sack_developmental_2016; @liu_scaling_2021] is not a developmental *fait accompli*. The lability of $D_\text{ad}$ may explain why there is so much putatively adaptive variation in the trait along light gradients [@muir_light_2018] and in coordination with other anatomical traits that vary among precipitation habitats [@pathare_increased_2020]. It also suggests that the relationship between genome size and stomatal anatomy at macroevolutionary scales [@roddy_scaling_2020] may not be causal. Genome size sets a minimum on meristematic cell volume [@simova_geometrical_2012], but the decoupling of size and density on the adaxial surface suggests that meristematic cell volume is probably not a strong constraint on the final size of pavement and guard cells.
